Strong lensing selection effects

### 2023-07-03 paper/seleff.tex
added sentence on color selection
resubmitted version

   \author{Alessandro Sonnenfeld\inst{\ref{sjtu},\ref{leiden}}\and
           Shun-Sheng Li\inst{\ref{leiden}}\and
           Giulia Despali\inst{\ref{ita}}\and
           Raphael Gavazzi\inst{\ref{marseille}}\and
           Anowar J. Shajib\inst{\ref{chicago},\ref{kicp},\ref{einstein}}\and
           Edward N. Taylor\inst{\ref{swin}}
          }

   \institute{
Department of Astronomy, School of Physics and Astronomy, Shanghai Jiao Tong University, Shanghai 200240, China\\
              \email{sonnenfeld@sjtu.edu.cn}\label{sjtu}
\and
Leiden Observatory, Leiden University, P.O. Box 9513, 2300 RA Leiden, The Netherlands\label{leiden}\and
Institut für Theoretische Astrophysik, Zentrum für Astronomie, Heidelberg Universität, Albert-Ueberle-Str. 2, 69120, Heidelberg, Germany\label{ita}\and
CNRS and CNES, Laboratoire d'Astrophysique de Marseille, Aix-Marseille Universit\'{e} 38, Rue Fr\'{e}d\'{e}ric Joliot-Curie, F-13388 Marseille, France\label{marseille}\and
Department of Astronomy \& Astrophysics, University of Chicago,
Chicago, IL 60637, USA\label{chicago}\and
Kavli Institute for Cosmological Physics, University of Chicago,
Chicago, IL 60637, USA\label{kicp}\and
NHFP Einstein Fellow\label{einstein}\and
Centre for Astrophysics and Supercomputing, Swinburne University of Technology, Hawthorn 3122, Australia\label{swin}
             }

@@

Strictly speaking, the results shown in \Sref{sect:results} apply only to samples of lenses and background sources with the same properties as our simulations.
{\bf Real lens samples can have more complex source distributions. For example, many existing lens-finding algorithms are tuned to identify sources with different colours from the lenses, introducing a selection that can modify the source distribution in redshift-magnitude space.
}
However, the comparison of section \ref{ssec:quasarbias} shows that, at fixed foreground galaxy population, the strong lensing bias for the sample of extended sources is indistinguishable from that of the lensed quasar population, at least for values of $\theta_{\mathrm{Ein,min}} < 1.0''$.
This result suggests that the dependence of the strong lensing bias on the details of the background source population is very weak, as long as no additional selection on the image configuration is applied (e.g. by selecting only quad lenses).
This is not a coincidence, but is a consequence of the fact that the dependence of the strong lensing cross-section on the lens parameters is a weak function of source magnitude\footnote{The cross-section itself is a strong function of source magnitude, but the trends between $\crosssect$ and the lens parameters are not.}, especially for magnitudes close to the detection limit (see section \ref{ssec:axisymmpoint}).
We can then conclude that the details of the properties of the background source population play a secondary role in determining the strong lensing bias.

@@

If one wishes to directly account for the strong lensing bias, the formally correct procedure is to explicitly model all of the selection steps in a Bayesian hierarchical formalism, as explained by \citet{Son22}.
Although this can be computationally challenging, machine learning can offer an efficient alternative \citep{Leg++22}.
In order for either of these approaches to work, however, it is essential that the lens selection procedure can be simulated.
This, in turn, requires having an objective definition of a strong lensing event. A peak-based definition such as that introduced in section \ref{ssec:lensdefext} could be used for this purpose.

@@

{\bf
Our extended source analysis was based on a peak detection criterion to define strong lensing events, as explained in section \ref{ssec:lensdefext}.
However, many previous studies aimed at simulating populations of lenses adopted a magnification-based definition, in which a lens-source system is classified as a strong lens if the magnification of the source is larger than a minimum value \citep[see e.g.][]{Col15, Rob++20}.
In order to check whether our results are robust to the choice of definition of a strong lensing event, we re-analysed the fiducial sample with this alternative definition. %repeated the analysis of the fiducial sample with extended sources. 
In particular, we defined as strong lenses all systems with a detected background source and $\mu_{\mathrm{tot}} > 3$, where $\mu_{\mathrm{tot}}$ is the total magnification (obtained by integrating the flux of all of the lensed images and dividing it by the intrinsic source flux).
The total number of strong lenses increased by $34\%$ with this new definition, mostly due to the inclusion of fainter sources with no detected multiple images.
\Fref{fig:minmubias} shows the strong lensing bias on a few selected quantities (stellar population synthesis mismatch parameter, projected dark matter mass within $5$~kpc, source magnitude and source half-light radius).
Besides the change in the source magnitude (third panel), we see very little differences in the bias with respect to the peak-based definition of the fiducial analysis. We then conclude that the exact definition of a strong lensing event does not have a big impact on the strong lensing bias of lens-related properties.
We stress out, however, that magnification is not a directly-observable quantity, and therefore we advise against applying magnification-based lens definitions to define lens samples, for the reasons discussed in section \ref{ssec:lensdefext}.
\begin{figure}
\includegraphics[width=\columnwidth]{minmu_bias.eps}
\caption{
Strong lensing bias in the case of a magnification-based lens definition (brown curves), compared to that of the fiducial analysis (blue curves).
First panel: median $\log{\asps}$.
Second panel: median $\mdmfive$ at $\log{\mstar}=11.5$ and $\log{\reff}=1.2$.
Third panel: median source magnitude.
Fourth panel: median half-light radius in the magnitude bin $24.8 < \msource < 25.2$.
}
\label{fig:minmubias}
\end{figure}
}

@@

In the time-delay study of \citet{Bir++20}, lenses are described with a mass density profile given by \Eref{eq:mstprofile}.
\citet{Bir++20} defined the profile by first fitting a pure power-law mass model to the strong lensing data, thus constraining the Einstein radius and power-law index $\gamma$, and then applying a mass-sheet transformation of the following kind:
\begin{equation}\label{eq:mst}
\kappa(\boldsymbol\theta) \rightarrow \lmst\kappa(\boldsymbol\theta) + 1 - \lmst.
\end{equation}
%\citet{Bir++20} fixed the slope $\gamma$ of the power-law component to the value obtained by fitting a pure power-law mass model to the strong lensing data.
The mass-sheet transformation parameter $\lmst$ can be interpreted as a quantity describing a departure from a pure power-law profile.
One possible way of estimating $\lmst$ for our lenses is to fit a power-law to the simulated images and then optimize for the value of $\lmst$ that best matches the true density profile. That, however, would be too time-consuming. Instead, we emulate this process as follows.

\citet{Son18} showed that, when the main arc and the counter-image of an extended source are well resolved, their relative widths constrain the following combination of derivatives of the lens potential at the Einstein radius:
\begin{equation}\label{eq:invariant}
\frac{\psi'''}{1 - \psi''}.
\end{equation}
In other words, the above quantity can be measured robustly (i.e. in a model-independent way).
When fitting a pure power-law lens model to lensing data, then, the value of $\gamma$ that we obtain is the one that reproduces the true $\psi'''/(1-\psi'')$\footnote{Strictly speaking, this is only true in the limit of well-resolved arcs and nearly axisymmetric lenses}. This is given by
\begin{equation}
\gamma = 2 + \tein \frac{\psi'''}{1-\psi''}
\end{equation}
Hence, given a lens in our sample, we first define a pure power-law model with the same Einstein radius as the true one and with $\gamma$ given by the equation above. 
Then, given a pure power-law model, we find $\lmst$ such that the first three derivatives of the lens potential match the true values.
In particular, since a mass-sheet transformation changes the second derivative of the potential as
\begin{equation}
\psi'' \rightarrow \lmst\psi'' + 1 - \lmst,
\end{equation}
we set
\begin{equation}
\lmst = \frac{1 - \psi''_{\mathrm{PL}}}{1 - \psi''},
\end{equation}
where $\psi''_{\mathrm{PL}}$ is the value of $\psi''$ of the pure power-law lens model.

>>>>

   \author{Alessandro Sonnenfeld\inst{\ref{sjtu},\ref{leiden}}\and
           Shun-Sheng Li\inst{\ref{leiden}}\and
           Giulia Despali\inst{\ref{bologna}}\and
           Raphael Gavazzi\inst{\ref{marseille}}\and
           Anowar J. Shajib\inst{\ref{chicago},\ref{kicp},\ref{einstein}}\and
           Edward N. Taylor\inst{\ref{swin}}
          }

   \institute{
Department of Astronomy, School of Physics and Astronomy, Shanghai Jiao Tong University, Shanghai 200240, China\\
              \email{sonnenfeld@sjtu.edu.cn}\label{sjtu}
\and
Leiden Observatory, Leiden University, P.O. Box 9513, 2300 RA Leiden, The Netherlands\label{leiden}\and
Alma Mater Studiorum - Università di Bologna, Dipartimento di Fisica e Astronomia "Augusto Righi", Via Gobetti 93/2, Bologna, Italy\label{bologna}\and
CNRS and CNES, Laboratoire d'Astrophysique de Marseille, Aix-Marseille Universit\'{e} 38, Rue Fr\'{e}d\'{e}ric Joliot-Curie, F-13388 Marseille, France\label{marseille}\and
Department of Astronomy \& Astrophysics, University of Chicago,
Chicago, IL 60637, USA\label{chicago}\and
Kavli Institute for Cosmological Physics, University of Chicago,
Chicago, IL 60637, USA\label{kicp}\and
NHFP Einstein Fellow\label{einstein}\and
Centre for Astrophysics and Supercomputing, Swinburne University of Technology, Hawthorn 3122, Australia\label{swin}
             }

@@

Strictly speaking, the results shown in \Sref{sect:results} apply only to samples of lenses and background sources with the same properties as our simulations.
{\bf Real lens samples can have more complex source distributions. For example, many existing lens-finding algorithms are tuned to identify sources with different colours from the lenses, introducing a selection that can modify the source distribution in redshift-magnitude space.
}
However, the comparison of section \ref{ssec:quasarbias} shows that, at fixed foreground galaxy population, the strong lensing bias for the sample of extended sources is indistinguishable from that of the lensed quasar population, at least for values of $\theta_{\mathrm{Ein,min}} < 1.0''$.
This result suggests that the dependence of the strong lensing bias on the details of the background source population is very weak, as long as no additional selection on the image configuration is applied (e.g. by selecting only quad lenses).
This is not a coincidence, but is a consequence of the fact that the dependence of the strong lensing cross-section on the lens parameters is a weak function of source magnitude\footnote{The cross-section itself is a strong function of source magnitude, but the trends between $\crosssect$ and the lens parameters are not.}, especially for magnitudes close to the detection limit (see section \ref{ssec:axisymmpoint}).
We can then conclude that the details of the properties of the background source population play a secondary role in determining the strong lensing bias.
{\bf 
A selection aimed at specific kinds of sources, such as blue star-forming galaxies or quasars, results primarily in different numbers of lenses being discovered, but does not change how biased the lenses are with respect to the general galaxy population.
}

@@

If one wishes to directly account for the strong lensing bias, the formally correct procedure is to explicitly model all of the selection steps in a Bayesian hierarchical formalism, as explained by \citet{Son22}.
Although this can be computationally challenging, machine learning can offer an efficient alternative \citep{Leg++23}.
In order for either of these approaches to work, however, it is essential that the lens selection procedure can be simulated.
This, in turn, requires having an objective definition of a strong lensing event. A peak-based definition such as that introduced in section \ref{ssec:lensdefext} could be used for this purpose.

@@

{\bf
Our extended source analysis was based on a peak detection criterion to define strong lensing events, as explained in section \ref{ssec:lensdefext}.
However, many previous studies aimed at simulating populations of lenses adopted a magnification-based definition, in which a lens-source system is classified as a strong lens if the magnification of the source is larger than a minimum value \citep[see e.g.][]{Col15, Robertson++20}.
In order to check whether our results are robust to the choice of definition of a strong lensing event, we re-analysed the fiducial sample with this alternative definition. %repeated the analysis of the fiducial sample with extended sources. 
In particular, we defined as strong lenses all systems with a detected background source and $\mu_{\mathrm{tot}} > 3$, where $\mu_{\mathrm{tot}}$ is the total magnification (obtained by integrating the flux of all of the lensed images and dividing it by the intrinsic source flux).
The total number of strong lenses increased by $34\%$ with this new definition, mostly due to the inclusion of fainter sources with no detected multiple images.
\Fref{fig:minmubias} shows the strong lensing bias on a few selected quantities (stellar population synthesis mismatch parameter, projected dark matter mass within $5$~kpc, source magnitude and source half-light radius).
Besides the change in the source magnitude (third panel), we see very little differences in the bias with respect to the peak-based definition of the fiducial analysis. We then conclude that the exact definition of a strong lensing event does not have a big impact on the strong lensing bias of lens-related properties.
We stress out, however, that magnification is not a directly-observable quantity, and therefore we advise against applying magnification-based lens definitions to define lens samples, for the reasons discussed in section \ref{ssec:lensdefext}.
\begin{figure}
\includegraphics[width=\columnwidth]{minmu_bias.eps}
\caption{
Strong lensing bias in the case of a magnification-based lens definition (brown curves), compared to that of the fiducial analysis (blue curves).
First panel: median $\log{\asps}$.
Second panel: median $\mdmfive$ at $\log{\mstar}=11.5$ and $\log{\reff}=1.2$.
Third panel: median source magnitude.
Fourth panel: median half-light radius in the magnitude bin $24.8 < \msource < 25.2$.
}
\label{fig:minmubias}
\end{figure}
}

@@

{\bf
In the time-delay study of \citet{Bir++20}, lenses are described with a mass density profile given by \Eref{eq:mstprofile}.
\citet{Bir++20} defined the profile by first fitting a pure power-law mass model to the strong lensing data, thus constraining the Einstein radius and power-law index $\gamma$, and then applying a mass-sheet transformation of the following kind:
\begin{equation}\label{eq:mst}
\kappa(\boldsymbol\theta) \rightarrow \lmst\kappa(\boldsymbol\theta) + 1 - \lmst.
\end{equation}
%\citet{Bir++20} fixed the slope $\gamma$ of the power-law component to the value obtained by fitting a pure power-law mass model to the strong lensing data.
The mass-sheet transformation parameter $\lmst$ can be interpreted as a quantity describing a departure from a pure power-law profile.
One possible way of estimating $\lmst$ for our lenses is to fit a power-law to the simulated images and then optimize for the value of $\lmst$ that best matches the true density profile. That, however, would be too time-consuming. Instead, we emulate this process as follows.

\citet{Son18} showed that, when the main arc and the counter-image of an extended source are well resolved, their relative widths constrain the following combination of derivatives of the lens potential at the Einstein radius:
\begin{equation}\label{eq:invariant}
\frac{\psi'''}{1 - \psi''}.
\end{equation}
In other words, the above quantity can be measured robustly (i.e. in a model-independent way).
When fitting a pure power-law lens model to lensing data, then, the value of $\gamma$ that we obtain is the one that reproduces the true $\psi'''/(1-\psi'')$\footnote{Strictly speaking, this is only true in the limit of well-resolved arcs and nearly axisymmetric lenses}. This is given by
\begin{equation}
\gamma = 2 + \tein \frac{\psi'''}{1-\psi''}
\end{equation}
Hence, given a lens in our sample, we first define a pure power-law model with the same Einstein radius as the true one and with $\gamma$ given by the equation above. 
Then, given a pure power-law model, we find $\lmst$ such that the first three derivatives of the lens potential match the true values.
In particular, since a mass-sheet transformation changes the second derivative of the potential as
\begin{equation}
\psi'' \rightarrow \lmst\psi'' + 1 - \lmst,
\end{equation}
we set
\begin{equation}
\lmst = \frac{1 - \psi''_{\mathrm{PL}}}{1 - \psi''},
\end{equation}
where $\psi''_{\mathrm{PL}}$ is the value of $\psi''$ of the pure power-law lens model.
}
